---
title: "Topic 5: k-Nearest Neighbors (k-NN)"
subtitle: "Distance-Based Classification and Model Selection"
author: "MS in Statistics for Data Science"
date: "UC3M, 2026"
output:
  html_document:
    theme: cosmo
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

# Introduction

In this notebook, we focus on **k-Nearest Neighbors (k-NN)**, a simple and powerful non-parametric classifier. We will study how distance metrics, scaling, and the choice of $k$ affect model behavior and performance.


```{r libraries}
# Core packages
library(tidyverse)    # Data manipulation and visualization
library(class)        # k-NN implementation
library(caret)        # Model training and evaluation
library(pROC)         # ROC curve analysis

# Visualization
library(ggplot2)
library(patchwork)
library(viridis)

# Set seed for reproducibility
set.seed(42)
```

---

# k-Nearest Neighbors for Classification

To practice with a classification example, we will use the iris dataset.

```{r iris-data}
# Load and explore the data
data(iris)
glimpse(iris)
```

```{r iris-visualization}
# Pairwise scatter plots
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Iris Dataset: Petal Dimensions",
       x = "Petal Length (cm)", y = "Petal Width (cm)") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```

```{r iris-split}
# Create train/test split
set.seed(123)
train_idx <- createDataPartition(iris$Species, p = 0.7, list = FALSE)
iris_train <- iris[train_idx, ]
iris_test <- iris[-train_idx, ]

cat("Training set:", nrow(iris_train), "samples\n")
cat("Test set:", nrow(iris_test), "samples\n")
```

k-NN uses distances, so features with larger scales dominate the distance calculation. Therefore, we should scale the data as a first step. Remember to create the scale parameters using only the train data

```{r scaling-importance}
# Without scaling - distances dominated by large features
cat("Range before scaling:\n")
sapply(iris_train[, 1:4], range)

# With scaling
train_scaled <- scale(iris_train[, 1:4])
test_scaled <- scale(iris_test[, 1:4],
                     center = attr(train_scaled, "scaled:center"),
                     scale = attr(train_scaled, "scaled:scale"))

cat("\nRange after scaling:\n")
sapply(as.data.frame(train_scaled), range)
```

## Using k-NN and decision boundaries

Once we have our scaled data, we can use the `knn` function.

```{r knn-basic}
# k-NN with k = 3
knn_pred_k3 <- knn(train = train_scaled,
                   test = test_scaled,
                   cl = iris_train$Species,
                   k = 3)

# Confusion matrix
confusionMatrix(knn_pred_k3, iris_test$Species)
```


To visualize the decision boundaries, we can create a grid of points using just 2 features, and get which class would be predicted at each element of the grid. This allows us to plot the ``backgroudn color''.

```{r knn-decision-boundary}
# Create a grid for visualization (using 2 features)
x_range <- seq(min(train_scaled[, "Petal.Length"]) - 0.5,
               max(train_scaled[, "Petal.Length"]) + 0.5, length.out = 100)
y_range <- seq(min(train_scaled[, "Petal.Width"]) - 0.5,
               max(train_scaled[, "Petal.Width"]) + 0.5, length.out = 100)
grid <- expand.grid(Petal.Length = x_range, Petal.Width = y_range)

# Predictions on grid for different values of k
grid$k1 <- knn(train = train_scaled[, c("Petal.Length", "Petal.Width")],
               test = as.matrix(grid[, c("Petal.Length", "Petal.Width")]),
               cl = iris_train$Species, k = 1)

grid$k5 <- knn(train = train_scaled[, c("Petal.Length", "Petal.Width")],
               test = as.matrix(grid[, c("Petal.Length", "Petal.Width")]),
               cl = iris_train$Species, k = 5)

grid$k15 <- knn(train = train_scaled[, c("Petal.Length", "Petal.Width")],
                test = as.matrix(grid[, c("Petal.Length", "Petal.Width")]),
                cl = iris_train$Species, k = 15)

# Plot decision boundaries
p_k1 <- ggplot() +
  geom_tile(data = grid, aes(x = Petal.Length, y = Petal.Width, fill = k1), alpha = 0.3) +
  geom_point(data = data.frame(train_scaled, Species = iris_train$Species),
             aes(x = Petal.Length, y = Petal.Width, color = Species), size = 2) +
  labs(title = "k = 1 (High Variance)") +
  theme_minimal() + theme(legend.position = "none")

p_k5 <- ggplot() +
  geom_tile(data = grid, aes(x = Petal.Length, y = Petal.Width, fill = k5), alpha = 0.3) +
  geom_point(data = data.frame(train_scaled, Species = iris_train$Species),
             aes(x = Petal.Length, y = Petal.Width, color = Species), size = 2) +
  labs(title = "k = 5 (Balanced)") +
  theme_minimal() + theme(legend.position = "none")

p_k15 <- ggplot() +
  geom_tile(data = grid, aes(x = Petal.Length, y = Petal.Width, fill = k15), alpha = 0.3) +
  geom_point(data = data.frame(train_scaled, Species = iris_train$Species),
             aes(x = Petal.Length, y = Petal.Width, color = Species), size = 2) +
  labs(title = "k = 15 (High Bias)") +
  theme_minimal() + theme(legend.position = "bottom")

(p_k1 | p_k5 | p_k15) +
  plot_annotation(title = "k-NN Decision Boundaries: Effect of k",
                  subtitle = "Small k = complex boundary (overfitting), Large k = smooth boundary (underfitting)")
```

## Selecting the Optimal k

The choice of k controls the bias-variance tradeoff:
- **Small k**: Low bias, high variance (complex, jagged boundaries)
- **Large k**: High bias, low variance (smooth boundaries)

```{r k-selection}
# Evaluate different values of k
k_values <- 1:30
accuracy <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  pred <- knn(train = train_scaled,
              test = test_scaled,
              cl = iris_train$Species,
              k = k_values[i])
  accuracy[i] <- mean(pred == iris_test$Species)
}

# Plot accuracy vs k
k_results <- data.frame(k = k_values, Accuracy = accuracy)

ggplot(k_results, aes(x = k, y = Accuracy)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_vline(xintercept = k_values[which.max(accuracy)],
             linetype = "dashed", color = "red") +
  annotate("text", x = k_values[which.max(accuracy)] + 2, y = max(accuracy),
           label = paste("Optimal k =", k_values[which.max(accuracy)]),
           color = "red") +
  labs(title = "k-NN: Selecting Optimal k",
       subtitle = "Test accuracy for different values of k",
       x = "Number of Neighbors (k)", y = "Accuracy") +
  theme_minimal()

cat("Optimal k:", k_values[which.max(accuracy)], "\n")
cat("Best accuracy:", max(accuracy), "\n")
```

We could make the $k$ value comparison more reboust by using cross validation with each possible $k$ value. This can be controlled with the `tuneGrid` option when using `caret::train()`.

```{r knn-caret}
# Set up cross-validation
ctrl <- trainControl(method = "cv", number = 10)

# Train k-NN with automatic k selection
knn_cv <- train(Species ~ .,
                data = iris_train,
                method = "knn",
                trControl = ctrl,
                preProcess = c("center", "scale"),
                tuneGrid = data.frame(k = 1:20))

# Results
print(knn_cv)
plot(knn_cv, main = "Cross-Validation Accuracy vs k")

# Best model predictions
knn_cv_pred <- predict(knn_cv, iris_test)
confusionMatrix(knn_cv_pred, iris_test$Species)
```

---

# k-Nearest Neighbors for Regression

k-NN can also predict continuous outcomes by averaging the target values of the nearest neighbors. Lets do an example where we predict `mpg` with `mtcars`

```{r knn-regression-data}
data(mtcars)

# Train/test split
set.seed(456)
train_idx_reg <- createDataPartition(mtcars$mpg, p = 0.75, list = FALSE)
mtcars_train <- mtcars[train_idx_reg, ]
mtcars_test <- mtcars[-train_idx_reg, ]

cat("Training set:", nrow(mtcars_train), "cars\n")
cat("Test set:", nrow(mtcars_test), "cars\n")
```

```{r knn-regression-fit}
# Cross-validation setup for regression
ctrl_reg <- trainControl(method = "repeatedcv", number = 5, repeats = 3)

# k-NN regression model
knn_reg <- train(
  mpg ~ wt + hp + disp + cyl,
  data = mtcars_train,
  method = "knn",
  trControl = ctrl_reg,
  preProcess = c("center", "scale"), # Scaling directly when training
  tuneGrid = data.frame(k = 1:15),
  metric = "RMSE"
)

print(knn_reg)
plot(knn_reg, main = "Cross-Validation RMSE vs k (Regression)")
```

We can get some regression metrics, and also plot the observed vs obtained results, which should fall in the red diagonal line if the fit was perfect.

```{r knn-regression-evaluation}
# Test set predictions
knn_reg_pred <- predict(knn_reg, newdata = mtcars_test)

# Regression metrics
rmse_val <- RMSE(knn_reg_pred, mtcars_test$mpg)
mae_val <- MAE(knn_reg_pred, mtcars_test$mpg)
r2_val <- R2(knn_reg_pred, mtcars_test$mpg)

cat("Best k:", knn_reg$bestTune$k, "\n")
cat("Test RMSE:", round(rmse_val, 3), "\n")
cat("Test MAE:", round(mae_val, 3), "\n")
cat("Test R^2:", round(r2_val, 3), "\n")

reg_results <- tibble(
  observed = mtcars_test$mpg,
  predicted = knn_reg_pred
)

ggplot(reg_results, aes(x = observed, y = predicted)) +
  geom_point(size = 3, color = "steelblue", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "k-NN Regression: Predicted vs Observed mpg",
       x = "Observed mpg",
       y = "Predicted mpg") +
  theme_minimal()
```

If we want to visualize the **shape** of the KNN regression function (like a step curve), we can do it with a model fitted in 1D, using only one predictor, here `wt`.

```{r knn-regression-shape}
# 1D model for visualization: mpg as a function of wt only
k_shape <- 3
knn_reg_wt <- train(
  mpg ~ wt,
  data = mtcars_train,
  method = "knn",
  trControl = ctrl_reg,
  preProcess = c("center", "scale"),
  tuneGrid = data.frame(k = k_shape),
  metric = "RMSE"
)

# Smooth grid over wt to draw the fitted function
wt_grid <- tibble(
  wt = seq(min(mtcars_train$wt), max(mtcars_train$wt), length.out = 300)
)

lm_wt <- lm(mpg ~ wt, data = mtcars_train)

plot_data <- wt_grid %>%
  mutate(
    mpg_knn = predict(knn_reg_wt, newdata = wt_grid),
    mpg_lm = predict(lm_wt, newdata = wt_grid)
  )

ggplot() +
  geom_point(data = mtcars_train, aes(x = wt, y = mpg),
             color = "firebrick", size = 2.2, alpha = 0.85) +
  geom_step(data = plot_data, aes(x = wt, y = mpg_knn),
            color = "royalblue", linewidth = 0.9) +
  geom_line(data = plot_data, aes(x = wt, y = mpg_lm),
            color = "black", linewidth = 1) +
  labs(
    title = "KNN Regression Shape for mpg vs wt",
    subtitle = paste("Blue: k-NN step fit (k =", k_shape, ") | Black: linear fit"),
    x = "wt (1000 lbs)",
    y = "mpg"
  ) +
  theme_minimal()
```

